---
title: "Loop Gotcha"
output: html_document
---

Recently I saw a tweet about a python behaviour I totally didn't expect. There's a very
strange interaction between list comprehensions and anonymous functions (lambdas). I later
realized the exact same thing happens in R's `for` loops. I've been using `*apply` and more
recently `purrr::map*` for long enough I forgot this behaviour existed. So let's see the
gotcha in action:

```{r}
funs <- list()
for(i in 1:10){
  funs[[i]] <- function(x) x^i
}

funs[[4]](2)
```

Hmmm, I was expecting this to return 16 (2^4), but instead I get 1024 (2^10).
It doesn't take long to see that this is because the value `i` ends up at 10,
and the function merely has a reference to the `i` that was defined in the loop.
So our `i` is mutating under the hood and our functions aren't doing what we wanted.

If that was the extent of the problem we'd be more or less ok. But the plot thickens.

```{r}
funs <- list()
for(i in 1:10){
  funs[[i]] <- function(x) x^i
}

for(i in 10:1)
  "Do Nothing"
  
funs[[4]](2)
```

What! Now I'm getting 2 (2^1), what gives? 

Well since the `i` escapes the loop, it is now sitting around in our global
environment for us to (accidentally) modify. After the second loop `i` is 
set to `1` and now all of our functions will return the number they're given.

Now we can fix this by creating a new scope for our loop body and re-binding
the index, then let the result escape the local scope.

```{r}
funs <- list()
for(i in 1:10)
  local({
    j <- i
    funs[[j]] <<- function(x) x^j
  })

funs[[4]](2)
```

or we could just do things the functional way:

```{r}
funs <- purrr::map(1:10, ~ function(x) x ^ .)
funs[[4]](2)
```

Much better `r emo::ji("wink")`
